---
title: "Evaluating Performance Using External Statistics with CDM"
output:
  pdf_document:
    number_sections: yes
    toc: yes
  html_document:
    number_sections: yes
    toc: yes
vignette: >
  %\VignetteIndexEntry{Using external evaluation with CDM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

\newpage
# Setup

Setup and data definitions
```{r setup, message=FALSE, warning=FALSE, eval=TRUE, results='hide'}
library(Eunomia)  # v1.0.2
library(FeatureExtraction)  # v3.2.0
library(PatientLevelPrediction)  # v6.0.4
library(glue)
library(plpDataAdapter)
library(LearningWithExternalStats)

# Activate Eunomia demo database
connectionDetails <- Eunomia::getEunomiaConnectionDetails()
Eunomia::createCohorts(connectionDetails)

# Data definitions
outcomeId <- 3
internalTargetId <- 1
externalTargetId <- 2

covSettings <- createCovariateSettings(
  useDemographicsGender = TRUE,
  useDemographicsAge = TRUE,
  useConditionGroupEraLongTerm = TRUE,
  useConditionGroupEraAnyTimePrior = TRUE,
  useDrugGroupEraLongTerm = TRUE,
  useDrugGroupEraAnyTimePrior = TRUE,
  useVisitConceptCountLongTerm = TRUE,
  longTermStartDays = -365,
  endDays = -1)

databaseDetails <- createDatabaseDetails(
  connectionDetails = connectionDetails,
  cdmDatabaseSchema = "main",
  cdmDatabaseName = '',
  cdmDatabaseId = '',  # ?
  tempEmulationSchema = NULL,  # is this important to avoid further errors in getPlpData?
  cohortDatabaseSchema = "main",
  cohortTable = "cohort",
  targetId = internalTargetId,
  outcomeDatabaseSchema = "main",
  outcomeTable = "cohort",
  outcomeIds = outcomeId,
  cdmVersion = 5  #?
)

restrictPlpDataSettings <- createRestrictPlpDataSettings()  # sampleSize = 1000
```
\newpage

# Internal node
We begin by defining prediction model settings:
```{r modelSettings, message=FALSE, warning=FALSE, eval=TRUE}
populationSettings <- createStudyPopulationSettings(
  washoutPeriod = 364,
  firstExposureOnly = FALSE,
  removeSubjectsWithPriorOutcome = TRUE,
  priorOutcomeLookback = 9999,
  riskWindowStart = 1,
  riskWindowEnd = 365,
  minTimeAtRisk = 364,
  requireTimeAtRisk = TRUE,
  includeAllOutcomes = TRUE)

splitSettings <- createDefaultSplitSetting(
  trainFraction = 0.75,
  testFraction = 0.25,
  type = 'stratified',
  nfold = 2,
  splitSeed = 1234
)

sampleSettings <- createSampleSettings()
featureEngineeringSettings <- createFeatureEngineeringSettings()
preprocessSettings <- createPreprocessSettings(
  minFraction = 0.01,
  normalize = T,
  removeRedundancy = T
)


lrModel <- setLassoLogisticRegression()

executeSettings = createExecuteSettings(
  runSplitData = T,
  runSampleData = T,
  runfeatureEngineering = T,
  runPreprocessData = T,
  runModelDevelopment = T,
  runCovariateSummary = T
)

```
Next we generate the internal data:
```{r generateInternal, message=FALSE, warning=FALSE, eval=TRUE, results='hide'}
internalPlpData <- getPlpData(
    databaseDetails = databaseDetails,
    covariateSettings = covSettings,
    restrictPlpDataSettings = restrictPlpDataSettings)
```
\newpage
We train the model and extract its active covariates:
```{r message=FALSE, warning=FALSE, eval=TRUE, results='hide'}
internalResults <- runPlp(
    plpData = internalPlpData,
    outcomeId = outcomeId,
    analysisId = '../singleDemo',
    analysisName = 'Demonstration of model training in internal node',
    populationSettings = populationSettings,
    splitSettings = splitSettings,
    sampleSettings = sampleSettings,
    featureEngineeringSettings = featureEngineeringSettings,
    preprocessSettings = preprocessSettings,
    modelSettings = lrModel,
    logSettings = createLogSettings(),
    executeSettings = executeSettings,
    saveDirectory = file.path(getwd(), '..', 'singlePlp')
  )
# Identify model covariates. These will be used in the estimation algorithm.
estimationCovariates <- getLRModelCovariates(internalResults)
```
Preliminary experiments have shown that using the subset selected by the model for generation of external statistics
gives better results than using all covariates.
```{r show results}
cat(glue('Selected {length(estimationCovariates)} features'), '\n')
```
The external node should receive now the estimation covariates.
\newpage

# External node
The external nodes generate PLP data and prepares statistics for sharing with the internal one. For the sake of 
algorithm evaluation it also evaluate the model trained on the internal node.

## Generate PLP data and evaluate model
```{r message=FALSE, warning=FALSE, eval=TRUE, results='hide'}
externalDatabaseDetails <- createDatabaseDetails(
  connectionDetails = connectionDetails,
  cdmDatabaseSchema = "main",
  cdmDatabaseName = '',
  cdmDatabaseId = '',  # ?
  tempEmulationSchema = NULL,  # is this important to avoid further errors in getPlpData?
  cohortDatabaseSchema = "main",
  cohortTable = "cohort",
  targetId = externalTargetId,
  outcomeDatabaseSchema = "main",
  outcomeTable = "cohort",
  outcomeIds = outcomeId,
  cdmVersion = 5  #?
)

externalPlpData <- getPlpData(
  databaseDetails = externalDatabaseDetails,
  covariateSettings = covSettings,
  restrictPlpDataSettings = restrictPlpDataSettings)

externalResults <- externalValidateDbPlp(  # TODO avoid duplicated generation of plpData?
  internalResults$model,
  validationDatabaseDetails = externalDatabaseDetails)
```

## Prepare statistics for sharing

```{r message=FALSE, warning=FALSE, eval=TRUE, results='hide'}
externalXY <- transformPlpDataToDataFrame(
  externalPlpData, populationSettings, outcomeId = outcomeId)

ZExt <- computeTable1LikeTransformation(
  externalXY[c(estimationCovariates, 'outcome')], 
  outcomeBalance = T, 
  outcomeCol = 'outcome'
)

muExt <- colMeans(ZExt)
```
\newpage

# Internal node: evaluation of external dataset peformence using internal dataset

```{r message=FALSE, warning=FALSE, eval=TRUE, results='hide'}
# Transform internal plp data and align it with predictions
internalXY <- transformPlpDataToDataFrame(
  internalPlpData, populationSettings, outcomeId = outcomeId)

allignedInternal <- allignDataAndPrediction(
  internalXY, internalResults, subsets = 'Train')

# Compute transformation on covariate outcome pairs
ZInt <- computeTable1LikeTransformation(
  allignedInternal$dataXY[c(estimationCovariates, 'outcome')], 
  outcomeBalance = T, 
  outcomeCol = 'outcome'
)

# Estimate external performance
internalData <- list(
  z=ZInt, 
  p = allignedInternal$prediction$value, 
  y = allignedInternal$dataXY[['outcome']]
)

estimatedResults <- estimateExternalPerformanceFromStats(
  internalData, 
  muExt, 
  divergence = 'entropy', lambda = 1e-2, minW = 1e-6, optimizationMethod = 'dual', 
  nboot = 10
)
```

# Compare evaluation and real results

```{r message=FALSE, warning=FALSE, eval=TRUE}
summarizeResults(
  internalResults$performanceEvaluation$evaluationStatistics, 'Test')

summarizeResults(
  externalResults[[2]]$performanceEvaluation$evaluationStatistics, 'Validation')

print(estimationSummaryToDF(estimatedResults$summary))
```
